Hydra Battles and Cie

### 2021-03-12 doc/chapter-powers.tex
consistent use of capitalization in headers

\subsection{Computing Fibonacci Numbers}

@@

\section{Representing Monoids in \coq \label{monoid-class-def}}

@@

\subsubsection{Multiplication on Peano Numbers}

@@

\subsubsection{String Concatenation}
We can use the notation \texttt{"\_ * \_ "} for other types than numbers.
In the following example,  the expression \texttt{"abc" * "def"} is interpreted
as \linebreak \texttt{@mult\_op string  {\color{darkred}?X} "abc"  "def"}, then the type  class mechanism replaces the unknown  {\color{darkred}?X} with 
\texttt{string\_op}.

@@

\subsubsection{Solving Ambiguities}
Let $A$ be some type, and let us assume there are several instances of
\texttt{Mult\_op $A$}. For solving ambiguity issues, one can
add a \emph{precedence} to each instance declaration of  
\texttt{Mult\_op $A$}. In any case, such ambiguity  can be addressed
by explicitly providing  some arguments of \texttt{mult\_op}.
For instance, in Sect.~\vref{nat-monoids}, we consider various monoids on types
\texttt{nat} and \texttt{N}.

\subsection{The Monoid Type Class}
\index{coq}{Type classes}
We are now ready to  give a definition of the \texttt{Monoid} class, using
\texttt{*} as an infix operator in scope \coqscope{M} for the monoid  multiplication.

@@

\subsection{Building Instances of \texttt{Monoid}}
Let \texttt{$A$} be some type, \texttt{{\it op}} an instance of 
\texttt{Mult\_op $A$} and \texttt{\it one: $A$}.
In order to build an instance of (\texttt{Monoid $A$ {\it op} {\it one}}),
one has to provide proofs of ``monoid axioms'' \texttt{ op\_assoc},
\texttt{one\_left} and \texttt{one\_right}.

@@

\subsection{Monoids and Equivalence Relations}
\index{coq}{Generalized rewriting}
\index{coq}{Type classes!Equivalence relations}

@@

\section{Computing Powers in any EMonoid}

@@

\subsection{The naive (linear) Algorithm}
The new version of the linear exponentiation function is as follows:

@@

\subsection{The Binary Exponentiation Algorithm}

@@

\subsection{Refinement and Correctness}
We have got two functions for computing powers in any monoid. 
So, it is interesting to ask oneself whether this duplication is useful, and which would be the respective role of \texttt{N\_bpow} and \texttt{power}.

@@

\section{Comparing Exponentiation Algorithms with Respect to Efficiency}

@@

\begin{Coqsrc}
Inductive computation {A:Type}  : Type :=
| Return (a : A)
| Mult (x y : A) (k : A -> computation).    
\end{Coqsrc}
\subsubsection{Monadic Notation}

@@

\subsubsection{Chain Correctness and Optimality}

@@

\section{Proving a Chain's Correctness}
\label{chain-correctness-sect}
In this section, we present various ways of proving that a given chain is 
correct w.r.t. a given exponent. First, we just try to apply 
the definition in Section~\vref{chain-correct-def}, but this method is very 
inefficient, even for small exponents. In a second step, we use more sophisticated techniques such as reflection and parametricity. Automatic generation of correct chains will be treated in Sect.~\vref{chain-generation}.

@@

\subsection{Correctness Proofs by Reflection}
\label{reflection-section}
\index{coq}{Proofs by reflection}
Instead of letting the tactic \texttt{rewrite} look for contexts in which
setoid rewriting is possible, we propose to use (deterministic) computations for
obtaining a ``canonical'' form for terms generated from a variable \texttt{x}
by contructors associated with monoid multiplication and neutral element.

@@

\section{Certified Chain Generators}
\label{chain-generation}

@@

\subsection{Chains and Continuations : f-chains}

>>>>

\subsection{Computing Fibonacci numbers}

@@

\section{Representing monoids in \coq \label{monoid-class-def}}

@@

\subsubsection{Multiplication on Peano numbers}

@@

\subsubsection{String concatenation}
We can use the notation \texttt{"\_ * \_ "} for other types than numbers.
In the following example,  the expression \texttt{"abc" * "def"} is interpreted
as \linebreak \texttt{@mult\_op string  {\color{darkred}?X} "abc"  "def"}, then the type  class mechanism replaces the unknown  {\color{darkred}?X} with 
\texttt{string\_op}.

@@

\subsubsection{Solving ambiguities}
Let $A$ be some type, and let us assume there are several instances of
\texttt{Mult\_op $A$}. For solving ambiguity issues, one can
add a \emph{precedence} to each instance declaration of  
\texttt{Mult\_op $A$}. In any case, such ambiguity  can be addressed
by explicitly providing  some arguments of \texttt{mult\_op}.
For instance, in Sect.~\vref{nat-monoids}, we consider various monoids on types
\texttt{nat} and \texttt{N}.

\subsection{The Monoid type class}
\index{coq}{Type classes}
We are now ready to  give a definition of the \texttt{Monoid} class, using
\texttt{*} as an infix operator in scope \coqscope{M} for the monoid  multiplication.

@@

\subsection{Building instances of \texttt{Monoid}}
Let \texttt{$A$} be some type, \texttt{{\it op}} an instance of 
\texttt{Mult\_op $A$} and \texttt{\it one: $A$}.
In order to build an instance of (\texttt{Monoid $A$ {\it op} {\it one}}),
one has to provide proofs of ``monoid axioms'' \texttt{ op\_assoc},
\texttt{one\_left} and \texttt{one\_right}.

@@

\subsection{Monoids and equivalence relations}
\index{coq}{Generalized rewriting}
\index{coq}{Type classes!Equivalence relations}

@@

\section{Computing powers in any EMonoid}

@@

\subsection{The naive (linear) algorithm}
The new version of the linear exponentiation function is as follows:

@@

\subsection{The binary exponentiation algorithm}

@@

\subsection{Refinement and correctness}
We have got two functions for computing powers in any monoid. 
So, it is interesting to ask oneself whether this duplication is useful, and which would be the respective role of \texttt{N\_bpow} and \texttt{power}.

@@

\section{Comparing exponentiation algorithms with respect to efficiency}

@@

\begin{Coqsrc}
Inductive computation {A:Type}  : Type :=
| Return (a : A)
| Mult (x y : A) (k : A -> computation).    
\end{Coqsrc}
\subsubsection{Monadic notation}

@@

\subsubsection{Chain correctness and optimality}

@@

\section{Proving a chain's correctness}
\label{chain-correctness-sect}
In this section, we present various ways of proving that a given chain is 
correct w.r.t. a given exponent. First, we just try to apply 
the definition in Section~\vref{chain-correct-def}, but this method is very 
inefficient, even for small exponents. In a second step, we use more sophisticated techniques such as reflection and parametricity. Automatic generation of correct chains will be treated in Sect.~\vref{chain-generation}.

@@

\subsection{Correctness proofs by reflection}
\label{reflection-section}
\index{coq}{Proofs by reflection}
Instead of letting the tactic \texttt{rewrite} look for contexts in which
setoid rewriting is possible, we propose to use (deterministic) computations for
obtaining a ``canonical'' form for terms generated from a variable \texttt{x}
by contructors associated with monoid multiplication and neutral element.

@@

\section{Certified chain generators}
\label{chain-generation}

@@

\subsection{Chains and continuations : f-chains}

### 2021-03-12 doc/chapter-primrec.tex
consistent use of capitalization in headers

\subsection{A little bit of Semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$ , together with the helper function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

@@

\subsubsection{Elementary Proofs of \texttt{isPR} statements}

@@

\subsubsection{Using Function Composition}

>>>>

\subsection{A little bit of semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$ , together with the helper function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

@@

\subsubsection{Elementary proofs of \texttt{isPR} statements}

@@

\subsubsection{Using function composition}

### 2021-03-12 doc/hydras.tex
renaming of a lemma
consistent use of capitalization in headers

\author{Pierre Castéran\\ LaBRI, Univ. Bordeaux, CNRS UMR 5800, \thanks{With contributions by Yves Bertot, \'Evelyne Contejean,  Florian Hatat and Pascal Manoury. \emph{The formalization of primitive recursive functions was originally authored by 
Russel O'Connor\cite{OConnor05}.}}}
\date{\today}
\title{Hydra Battles and Cie}
\makeindex{coq}
\makeindex{maths}
\makeindex{hydras}
\makeindex{additions}
\makeindex{primrec}
\begin{document}
\maketitle

@@

@@

\subsection{Trust in our Proofs}
\label{sect:trust-in-proofs}

@@

\subsection{Assumed Redundancy}

It may often happen that several definitions of a given concept, or several proofs of a given theorem are possible. If all the versions present some interest, we will make them available, since each one may be of some methodological 
interest (by illustrating some tactic of proof pattern, for instance).
We may use \coq's module system to make several proofs of a given theorem co-exist in our libraries (see also Sect~\vref{sect:alt-proofs}).
After some discussions of the pros and cons of each solution, we develop only one of them, leaving the others  as exercises or projects (i.e., big or difficult exercises).
In order to discuss which assumptions are really needed for proving a theorem, we will also present 
several aborted proofs.
Of course, do not hesitate to contribute nice proofs or alternative definitions !

It may also happen that some proof looks to be useless, because the proven theorem is a trivial consequence of another (proven too) result.
For instance, let us consider the three following statements:
\begin{enumerate}
\item There is no measure into $\mathbb{N}$ for proving the termination of all hydra battles (Sect~\vref{omega-case}).
\item There is no measure into $[0,\omega^2)$ for proving the termination of all hydra battle (Sect~\vref{omega2-case}).
\item There is no measure into $[0,\mu)$ for proving the termination of all hydra battles, for any $\mu<\epsilon_0$ (Sect\vref{sec:free-battles-case}).
\end{enumerate}

@@

\subsection{About Logic}

@@

\subsection{Alternative or Bad Definitions}
\label{sect:alt-proofs}
Finally, we decided to include definitions or lemma statements, as well as tactics,  that lead to
dead-ends or too complex developments, with the following coloring.
Bad definitions 
 and encapsulation in modules called \texttt{Bad}, \texttt{Bad1}, etc.

@@

\section{How to Install the Libraries}
\label{sec:orgheadline4}
\begin{itemize}
\item The present distribution has been checked with version 8.13.0 of the Coq proof assistant, with the plug-ins \texttt{coq-paramcoq}, \texttt{coq-equations}
and \texttt{coq-mathcomp-algebra}.

\item Just go into the root directory, and type "\texttt{make}" ("\texttt{make -k}"  in case of an incomplete installation).
\end{itemize}

\section{Comments on Exercises and Projects}

@@

\include{part-hydras}
\include{chapter-primrec}
\part{Some  Certified Algorithms}

>>>>

\author{Pierre Castéran\\ LaBRI, Univ. Bordeaux, CNRS UMR 5800, \thanks{With contributions by Yves Bertot, \'Evelyne Contejean,  Florian Hatat and Pascal Manoury. The formalization of primitive recursive functions was originally authored by 
Russel O'Connor\cite{OConnor05}.}}
\date{\today}
\title{Hydra Battles and Cie}
\makeindex{coq}
\makeindex{maths}
\makeindex{hydras}
\makeindex{additions}
\makeindex{primrec}
\begin{document}
\maketitle

@@

Let us underline the analogy between hydra battles and interactive theorem proving. Hercules is the user (you!), and hydra's heads are the sub-goals: you may think that applying a tactic would solve a sub-goal, but it results often in the multiplication of such tasks.

@@

\subsection{Trust in our proofs}
\label{sect:trust-in-proofs}

@@

\subsection{Assumed redundancy}

It may often happen that several definitions of a given concept, or several proofs of a given theorem are possible. If all the versions present some interest, we will make them available, since each one may be of some methodological 
interest (by illustrating some tactic of proof pattern, for instance).
We use \coq's module system to make several proofs of a given theorem co-exist in our libraries (see also Sect~\vref{sect:alt-proofs}).
After some discussions of the pros and cons of each solution, we develop only one of them, leaving the others  as exercises or projects (i.e., big or difficult exercises).
In order to discuss which assumptions are really needed for proving a theorem, we will also present 
several aborted proofs.
Of course, do not hesitate to contribute nice proofs or alternative definitions!

It may also happen that some proof looks to be useless, because the proven theorem is a trivial consequence of another (proven too) result.
For instance, let us consider the three following statements:
\begin{enumerate}
\item There is no measure into $\mathbb{N}$ for proving the termination of all hydra battles (Sect~\vref{omega-case}).
\item There is no measure into the interval\footnote{We use the notation $[a,b)$ for denoting the set of ordinals greater or equal than $a$ and strictly less than $b$.} interval $\{x\$[0,\omega^2)$ for proving the termination of all hydra battles (Sect~\vref{omega2-case}).
\item There is no measure into $[0,\mu)$ for proving the termination of all hydra battles, for any $\mu<\epsilon_0$ (Sect\vref{sec:free-battles-case}).
\end{enumerate}

@@

\subsection{About logic}

@@

\subsection{Alternative or bad definitions}
\label{sect:alt-proofs}
Finally, we decided to include definitions or lemma statements, as well as tactics,  that lead to
dead-ends or too complex developments, with the following coloring.
Bad definitions 
 and encapsulation in modules called \texttt{Bad}, \texttt{Bad1}, etc.

@@

\section{How to install the libraries}
\label{sec:orgheadline4}
\begin{itemize}
\item The present distribution has been checked with version 8.13.0 of the Coq proof assistant, with the plug-ins \texttt{coq-paramcoq}, \texttt{coq-equations}
and \texttt{coq-mathcomp-algebra}.

\item Please refer to \href{https://github.com/coq-community/hydra-battles#readme}{the README file of the project}
\end{itemize}

\section{Comments on exercises and projects}

@@

\include{part-hydras}
\include{chapter-primrec}
\part{A Few  Certified Algorithms}
